---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment


In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv("data/county.csv", header = TRUE, sep = ",")
county_MI <- read.csv("data/county_MI.csv", header = TRUE, sep = ",")
```

* What does each row represent?

# Each row represents a county within a certain state.


Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

* Are there any NA values? 

# County_MI.csv did not have any NA values. Income, Income_Err, and Child Poverty return with 1 NA each.

* What differences do you notice between the structure of the two data frames?

# The file county_MI has one more variable than county_df.

```{r}
summary(county_df)
summary(county_MI)
str(county_df)
str(county_MI)
```

The county_MI data frame contains data for Michigan only and includes a
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for median Income

Note that the field called `Income` is the median income
for that county. You do **NOT** need to compute a median. The values are the medians.

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 
Write a short paragraph describing the distribution of median incomes.

Median income is centered around approximately $42,000.  It is skewed to the right, where there are very few people earning more than $75,000, relative to the population. Based on the violin plot, the majority of the population lies between $30,000 - $60,000.
```{r}
ggplot(county_df) + geom_histogram(aes(x=Income), fill="red", colour="black")
ggplot(county_df) + geom_boxplot(aes(x="Boxplot of Median Household Income Across America", y=Income))
ggplot(county_df) + geom_violin(aes(x="Boxplot of Median Household Income Across America", y=Income))
ggplot(county_df) + geom_density(aes(x=Income))



```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US. 

```{r histo_totpop}
ggplot(county_df) + geom_histogram(aes(x=TotalPop), fill="#884444", colour="blue")
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
ggplot(county_df) + 
  geom_histogram(aes(x=TotalPop), 
              fill="gray", colour="purple") +
  ggtitle("Distribution of population for US counties") + xlab("Total Population") + ylab("Number of counties")
```

Repeat the histogram but only for counties with less than 1.5 million people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m}
ggplot(county_df[county_df$TotalPop <= 1500000,]) + geom_histogram(aes(x=TotalPop), fill = "gray", colour = "purple") + ggtitle("Distribution of population for US counties") + xlab("Total Population") + ylab("Number of counties")
```


**Hacker Extra** - Create a function that uses the
[Freedman-Diaconis](https://en.wikipedia.org/wiki/Freedman%E2%80%93Diaconis_rule) rule for calculating bin width and number of bins. The function should take a vector (e.g. a dataframe column) as its
input and return the integer number of bins recommended for the histogram.

```{r}
numbins_FD <- function(data_col){
  
}
```



### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped violin plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?

The peninsulas have a relatively similar median point, around $40,000 give or take.  The UP definitely has fewer extremes, everyone in the UP is pretty much on the same field.

```{r vp_inc_by_peninsula}
ggplot(county_MI) + geom_violin(aes(x=peninsula, y=Income), fill = "gray", colour = "purple", trim = FALSE) + ggtitle("Income by Peninsula") + xlab("Peninsula") + ylab("Income")


```

Create several boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

The LP has a far higher median income than the UP, with that there is more variation in the LP.  It is so close it seems like you could say the 1st quartile of the LP meets the 3rd quartile of the UP.

Once again we see high variation in the LP and a relatively tight group in the UP.  It is interesting to see that the boxes are not too far off from each other 

For some reason the third one surprised me as I imagined people living in the UP having to drive far to "get into town" so to speak.  But it would seem that the opposite is true.  I assume this is because the communities in the UP all live in smaller towns and only travel far when need be.  On the other hand, I know many people who drive an hour or more each day to get down to Detroit.  I'm a bad guesser I suppose.

I was again surpirsed by the fourth one with there being so many professionals in the UP compared to the LP.  This is evident by the mass of the box along the line.  Of course, the LP has higher outliers on this mark, but that was a give away.

From the fifth graph, I can conclude that the UP has a denser population of those that descend from the Angol-Saxon heritage as compared to the LP.

From the sixth graph, I can conclude that the UP has a much denser population of those that descend from the Native American heritage as compared to the LP.

```{r boxplots}

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=Income), fill = "gray", colour = "purple") + ggtitle("Income by Peninsula") + xlab("Peninsula") + ylab("Income")

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=Poverty), fill = "gray", colour = "purple") + ggtitle("Poverty by Peninsula") + xlab("Peninsula") + ylab("Poverty")

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=MeanCommute), fill = "gray", colour = "purple") + ggtitle("Mean Commute by Peninsula") + xlab("Peninsula") + ylab("Mean Commute")

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=Professional), fill = "gray", colour = "purple") + ggtitle("Professional Job by Peninsula") + xlab("Peninsula") + ylab("Professional")

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=White), fill = "gray", colour = "purple") + ggtitle("White Ancestry by Peninsula") + xlab("Peninsula") + ylab("White Ancestry")

ggplot(county_MI) + geom_boxplot(aes(x=peninsula, y=Native), fill = "gray", colour = "purple") + ggtitle("Native Ancestry by Peninsula") + xlab("Peninsula") + ylab("Native")
```

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how unemployment rate compares by region.  Make sure your graph has a title and good axis
labels. 

```{r faceted_histo}
ggplot(county_df) + geom_histogram(aes(x=Unemployment), fill = "gray", colour = "purple") + ggtitle("Unemployment by Region") + xlab("Unemployment") + ylab("Number of Counties") + facet_wrap(~Region)

```

Now do a group of boxplots of mean commute time by region.

```{r bp_commute_region}
ggplot(county_df) + geom_boxplot(aes(x=Region, y=MeanCommute), fill = "gray", colour = "purple") + ggtitle("Mean Commute by Region") + xlab("Region") + ylab("Mean Commute Time")

```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Division in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_division}
ggplot(county_df) + geom_bar(aes(x = Division), fill = "gray", colour = "purple") + ggtitle("US Counties by Division") + xlab("Division") + ylab("County") + coord_flip()

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(county_MI) + geom_point(aes(x=Income, y=MeanCommute), fill = "gray", colour = "purple") + ggtitle("Michigan Income by Mean Commute Time") + xlab("Income") + ylab("Mean Commute Time")
```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population.

```{r xy_inc_commute_peninsula}
ggplot(county_MI) + geom_point(aes(x=Income, y=MeanCommute, colour=peninsula,size=TotalPop)) + ggtitle("Michigan Income by Mean Commute Time") + xlab("Income") + ylab("Mean Commute Time")
```

Your thoughts on the relationship between these variables?

# Based off this one chart, one might be inclined to conclude that longer commute times lead to higher incomes, but outside of this chart it remains to be seen whether or not this is truly the case.

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

The first scatter plot shows more clearly that increase unemployment levels may have led to increased poverty levels.  The UP displays a generally flat horizontal line for the same criteria.

The second chart seems to show a slight relation in the LP of production to poverty levels.  Again, the UP is a flat line; I assume this is beause of the small data size.

```{r xy_faceted_peninsula}
ggplot(county_MI) + geom_point(aes(x=Unemployment, y=Poverty, colour=peninsula)) + ggtitle("Michigan Unemployment by Poverty") + xlab("Unemployment") + ylab("Poverty") + facet_wrap(~peninsula)

ggplot(county_MI) + geom_point(aes(x=Production, y=Poverty, colour=peninsula)) + ggtitle("Michigan Production to Poverty") + xlab("Production") + ylab("Poverty") + facet_wrap(~peninsula)

```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

There seems to be correlation between the income and population size, which fits the narrative of more rural areas having lower wages than in more densly populated areas like cities.

```{r xy_pop_inc}
ggplot(county_df) + geom_point(aes(x=TotalPop, y=Income), colour="purple") + ggtitle("Michigan Total Population to Income") + xlab("Total Population") + ylab("Income") + scale_x_log10()
```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.